```{r, include=FALSE}
# at some point this should be moved into bookdown
source("~/Desktop/hpcbio-git/caceres-16S-mosquito-2022Sept/src/common_code.R", local = knitr::knit_global())
```

```{r}
physeq.filtered <- readRDS('results/physeq.filtered.RDS')
```

# Alpha diversity

Check alpha diversity stats.  Note we haven't performed any additional filtering or agglomeration at this stage; this is largely due to recommendations to prevent loss of rare ASVs or singletons in the data, which can dramatically affect alpha diversity measures.  Specifically (from the `plot_richness` function used here):

> You must use untrimmed, non-normalized count data for meaningful results, as many of these estimates are highly dependent on the number of singletons. You can always trim the data later on if needed, just not before using this function.

## Normalize and calculate alpha diversity 


We'll calculate some initial alpha diversity metrics.

```{r echo=FALSE, message=FALSE}
physeq.tmp <- prune_samples(sample_data(physeq.filtered)$Density != "Stock Water", physeq.filtered)
# all six measures: Observed, Chao1, ACE, Shannon, Simpson, Inv. Simpson, Fisher
erDF <- estimate_richness(physeq.tmp, 
                          measures = c('Observed', 'Chao1', 'Shannon', 'Simpson', 'InvSimpson'))

# add Faith's PD, this is a simple wrapper to format everything correctly
pd <- estimate_pd(physeq.tmp)

# we don't need SR, it's the same as observed
erDF$PD <- pd$PD

measures <- colnames(erDF)
```

We'll check the overall measures for alpha diversity across different groups.  The descriptions below are largely from the [QIIME2 forum](https://forum.qiime2.org/t/alpha-and-beta-diversity-explanations-and-commands/2282) with some additional notes.

* **Observed** - Number of features (ASVs in this case).  From the `phyloseq` package, function `estimate_richness`. 
* **Chao1** - Estimates number of rare taxa missed from undersampling, with an associated confidence interval. Compare to observed taxa, the difference will be those expected to be missing. From the `phyloseq` package, function `estimate_richness`. 
* **ACE** - Abundance-based Coverage Estimator. Estimates species richness using a correction factor.  From the `phyloseq` package, function `estimate_richness`. 
* **Shannon** - Calculates richness and diversity using a natural logarithm, accounts for both abundance and evenness of the taxa present. From the `vegan` package, function `diversity`. 
* **Simpson** - Measures the relative abundance of the different species making up the sample richness.  From the `vegan` package, function `diversity`. 
* **Inverse Simpson** - Measures the relative abundance of the different species making up the sample richness.  From the `vegan` package, function `diversity`. 
* **Fisher** - Relationship between the number of species and the abundance of each species.  From the `vegan` package, function `diversity`. 
* **Faith's PD** - Sum of the total phylogenetic branch length for one or multiple samples.  From the `picante` package, function `pd`. 

## Check for normal distribution 

### Plot to check normality

Do these look like a normal distribution? If p is less than 0.05, then it is not normally distributed and non-parametric tests should be used instead (such as the Wilcoxen-rank sum test).

```{r }
par(mfrow = c(2, 3))

plots <- lapply(c('Chao1', 'Shannon', 'Simpson', 'InvSimpson', 'Observed', 'PD'), 
       function(x) {
         shap <- shapiro.test(erDF[,x])
         hist(erDF[,x], xlab = "Measure", main=paste0(x, "\nShapiro pvalue=", label_number(accuracy = 0.0001)(shap$p.value)), breaks=15)
         })
```

Chao1, Observed, & Faith PD are normal. The others are skewed. Using non-parametric tests in the visualizations below. Parametric tests were causing errors, but I can revisit them.

Look at significance of metrics across Env_Comp_Den using Kruskal Wallace test (non-parametric ANOVA). Note, Stock Water was removed. The Kruskal Wallace test would not work with it present.

```{r}
p <- plot_richness_estimates(physeq = physeq.tmp, 
                             erDF = erDF[c(1,2,5,7)],
                             "Env_Comp_Den", 
                             color = "Simple_extraction", 
                             title="Alpha diversity of Env_Comp_Den (Kruskal Wallace test)")

p + geom_boxplot(aes(x=Env_Comp_Den,
                     y=value,
                     color=NULL), alpha=0.1) +
  stat_compare_means(label = "p.signif",
                #     ref.group = "Infusion",
                     hide.ns = TRUE,
                     method = "kruskal.test") +
  ylab("Diversity value") 
  #scale_x_discrete(labels=labels) + 
  #theme(legend.position = "none")

#ggsave("alpha-allMetrics-byEnv_Comp_Den.pdf", path = "./results/final-plots", device = "pdf", width = 8, height = 5.5, units = "in")
```

In the metrics that consider abundance and richness (Shannon, Simpson, InvSimpson), all groups are significant. This essentially means that the alpha diversity values of one of these groups compared to the mean or any other group were significantly different. Technically the other metrics had a normal distribution and would be better suited for an ANOVA test. This can be added later.


Also looked at significance of metrics for Density using Kruskal Wallace test (non-parametric ANOVA). Note, Stock Water was removed. The Kruskal Wallace test would not work with it present.

```{r}
p <- plot_richness_estimates(physeq = physeq.tmp, 
                             erDF = erDF[c(1,2,5,7)],
                             "Density", 
                             color = "Simple_extraction", 
                             title="Alpha diversity of Density (Kruskal Wallace test)")

p + geom_boxplot(aes(x=Density,
                     y=value,
                     color=NULL), alpha=0.1) +
  stat_compare_means(label = "p.signif",
                #     ref.group = "Infusion",
                     hide.ns = TRUE,
                     method = "kruskal.test") +
  ylab("Diversity value") 
  #scale_x_discrete(labels=labels) + 
  #theme(legend.position = "none")

#ggsave("alpha-allMetrics-byDensity.pdf", path = "./results/final-plots", device = "pdf", width = 7, height = 5.5, units = "in")
```
