```{r, include=FALSE}
source("~/Desktop/hpcbio-git/caceres-16S-mosquito-2022Sept/src/common_code.R", local = knitr::knit_global())
```

```{r}
physeq.prev <- readRDS('../results/phyloseq.prevfiltered.RDS')
```

# Beta diversity

Beta diversity tells us how much diversity pairwise samples share with one another. Beta diversity metrics assign a value to that shared diversity. Bray-Curtis is a common non-phylogenetic metric used to compare sample composition based on presence and abundance of ASVs. Weighted Unifrac is a common phylogenetic metric that also compares phylogenetic distance of the shared diversity between two samples. If you have a preferred metric other than these, I can add them in. 

To plot these measurements we have to choose an appropriate ordination method. PCA is NOT recommended for microbial data, since it is not linear, unless it is Hellinger-transformed. PCoA and NMDS are recommended for unconstrained ordination (axes are not constrained by sample metadata), while the former is used in most circumstances. CA is recommended for constrained ordination, however rare data could have an unduly large influence on CA. See [Qian et al., 2020](`https://journals.lww.com/cmj/Fulltext/2020/08050/A_guide_to_human_microbiome_research__study.10.aspx`) for more details. For these reasons, I'll stick with PCoA unless requested otherwise.

## Heatmaps
Below is an initial beta diversity heatmap based on the prevalence filtered and taxa-glommed samples at the Family level (y-axis). This is based on PCoA values (recommended for 16S data) and the Bray-curtis (non-phylogenetic) metric. Data is log2 transformed. Note that this does show all 155 ASVs on the y-axis, which means there is some redundancy in Family taxa on that axis. 

```{r }
p <- plot_heatmap(physeq.prev,
                  method = "PCoA",
                  distance = "bray",
                  taxa.label = "Family",
                  sample.label = "Label",
                  trans = log_trans(2),
                  title = "PCoA Bray-Curtis Heatmap")
p
```

This heatmap looks very promising. All G's are on the right, and all GN's are on the left. These further separate by water versus mosquito larvae. Let's also try it using a phylogenetic metric, weighted unifrac.

```{r }
p <- plot_heatmap(physeq.prev,
                  method = "PCoA",
                  distance = "wunifrac",
                  taxa.label = "Family",
                  sample.label = "Label",
                  trans = log_trans(2),
                  title = "PCoA Weighted Unifrac Heatmap")
p
```
This looks the same, which tells us this pattern is robust to differing metrics.

Here I have made the same kind of heatmap but with the data glommed at the Family level, so there is no redundancy on the y-axis. However, this does have the downside of grouping all NA taxa into the same group, which is not ideal either, so I've labeled the NA taxa by their Order instead.

```{r}
#Agglomerate taxa and species level to remove redundancy
physeq.heat.glom <- tax_glom(physeq.prev, taxrank = "Family", NArm = FALSE)

# Family is not NA
no.na <- !is.na(tax_table(physeq.heat.glom)[,"Family"])

#Family level in beneficial has a couple NAs that need to be renamed to Family
tax_table(physeq.heat.glom)[!no.na][,"Family"] = paste("[Order]", 
                                                      tax_table(physeq.heat.glom)[!no.na][,"Order"])
```

```{r }
p <- plot_heatmap(physeq.heat.glom,
                  method = "PCoA",
                  distance = "bray",
                  taxa.label = "Family",
                  sample.label = "Label",
                  trans = log_trans(2),
                  title = "PCoA Bray-Curtis Heatmap - Glommed at Family Level")
p
```
This still separates by water vs larvae, but it no longer separates as nicely by environment, so I would not recommend using the above plot.


## Normalization & PCoA (mosquito & stock water)

Transform data to relative proportions (no prior added).  While not as sophisticated as other normalization methods it is currently the best method for normalizing library sizes while still maintaining community structure patterns. See [McKnight et al., 2018](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.13115) for more information.

```{r }
#Remove experimental water samples
physeq.nowat <- prune_samples(sample_data(physeq.prev)$Simple_extraction != "Water", physeq.prev)

#Normalize
physeq.prop.nowat <- transform_sample_counts(physeq.nowat, function(x) x/sum(x) )
sample_data(physeq.prop.nowat)$SampleSums <- sample_sums(physeq.nowat)
```

Set the active normalization type (proportion for now)

```{r}
physeq.active <- physeq.prop.nowat
```

### Bray Curtis metric



```{r }
set.seed(1234)

finalPhyseq.ord <- ordinate(physeq.active, "PCoA", "bray")
p1 = plot_ordination(physeq.active,
                     finalPhyseq.ord,
                     type="samples",
                     color="Environment",
                     shape = "Simple_extraction",
                     title="Samples") + geom_point(size = 2.5)
p1 + ggtitle("PCoA, Bray Curtis distance")
 # facet_wrap(~Simple_extraction)
#ggplotly(p1 + ggtitle("PCoA, Bray Curtis distance"))
#+  scale_fill_viridis_d()
```

Very clearly splitting by Environment and extraction type (simplified to water vs larvae). Larvae samples also cluster more tightly than water samples, which isn't too surprising.

Let's try highlighting Density and Competition...

```{r }
set.seed(1234)

finalPhyseq.ord <- ordinate(physeq.active, "PCoA", "bray")
p1 = plot_ordination(physeq.active,
                     finalPhyseq.ord,
                     type="samples",
                     color="Environment",
                     shape = "Density",
 #                   label = "Environment",
                     title="PCoA, Bray Curtis distance") + 
  geom_point(size = 3, aes(fill=Competition)) + 
  scale_shape_manual(values = c(21, 22, 23) )+
  scale_color_manual(values = c("black","gray") ) +
  scale_fill_manual(values=c("#EC9DAE","#AEDE94", "#9dd6ec")) +
  #scale_fill_manual(values=c("black","darkgray", "white")) +
  guides(fill=guide_legend(override.aes=list(shape=21)))+
  guides(color=guide_legend(override.aes=list(shape=21))) 
  guides(fill=guide_legend(override.aes=list(shape=21,fill=c("#EC9DAE","#AEDE94","#9dd6ec"),color=c("black","white"))))
  #guides(fill=guide_legend(override.aes=list(shape=21,fill=c("black","darkgray", "white"),color=c("black","gray"))))
p1
ggsave("pcoa-braycurtis-density-competition-environment_Mosquito_color.pdf", path = "../results/final-plots/", device = "pdf")

#aes(alpha=Environment)
#ifelse(, NA, Competition)
#aes(color=Density, shape=Competition, fill=Env_bool),
#scale_shape_manual(values=c(21,22,23)) +
#  scale_fill_discrete(na.value=NA, guide="none")
```

One could argue that Density 40 are slighty more skewed towards the right and Density 20 to the left.


Check to see if number of sequences per sample (input) affects diversity

```{r }
p1 <- plot_ordination(physeq.active,
                     finalPhyseq.ord,
                     type="samples",
                     color="input",
                     label="Label",
                     title="Samples") + geom_point(size = 2.5) +
  ggtitle("PCoA, Bray Curtis distance") 
ggplotly(p1)
```

Nothing odd related to number of input sequences.


### Weighted UniFrac metric

Let's look at Weighted UniFrac now.

```{r}
finalPhyseq.ord <- ordinate(physeq.active, "PCoA", "wunifrac")
p1 = plot_ordination(physeq.active,
                     finalPhyseq.ord,
                     type="samples",
                     color="Environment",
                     shape = "Simple_extraction",
                    # label = "Label",
                     title="PCoA, Weighted Unifrac distance") + geom_point(size = 2.5)
p1

```

Very similar pattern as Bray-Curtis but not quite as strong. I bit more mixing between G & GN samples. Since weighted unifrac takes phylogenetic distance into account this demonstrates that the phylogenetic distance of the ASVs (and potentially functional similarity) is a bit more close for some of the G/GN samples. There also appear to be 3 outlier mosquito larvae that are different from the main group, but this is based on phylogeny because this pattern isn't seen in Bray Curtis.

Let's also highlight the other metadata of interest:

```{r}
finalPhyseq.ord <- ordinate(physeq.active, "PCoA", "wunifrac")
p1 = plot_ordination(physeq.active,
                     finalPhyseq.ord,
                     type="samples",
                     color="Environment",
                     shape = "Density",
                     title="PCoA, Weighted Unifrac distance") +
  geom_point(size = 3, aes(fill=Competition)) + 
  scale_shape_manual(values = c(21, 22, 23) )+
  scale_color_manual(values = c("black","gray") ) +
  #scale_fill_manual(values=c("#EC9DAE","#AEDE94", "#9dd6ec")) +
  scale_fill_manual(values=c("black","darkgray", "white")) +
  guides(fill=guide_legend(override.aes=list(shape=21)))+
  guides(color=guide_legend(override.aes=list(shape=21))) 
  #guides(fill=guide_legend(override.aes=list(shape=21,fill=c("#EC9DAE","#AEDE94","#9dd6ec"),color=c("black","white"))))
  guides(fill=guide_legend(override.aes=list(shape=21,fill=c("black","darkgray", "white"),color=c("black","gray"))))
  
p1

ggsave("pcoa-wunifrac-density-competition-environment_Mosquito_bgw.pdf", path = "../results/final-plots/", device = "pdf")
```

Again, one could argue that Density splits somewhat, but pretty well mixed.



## PCoA (experimental water & stock water) 

I will just look at experimental water and stock water now.

```{r }
#Remove experimental water samples
physeq.wat <- prune_samples(sample_data(physeq.prev)$Simple_extraction != "Mosquito Larvae", physeq.prev)

#Normalize
physeq.prop.wat <- transform_sample_counts(physeq.wat, function(x) x/sum(x) )
sample_data(physeq.prop.wat)$SampleSums <- sample_sums(physeq.wat)
```


```{r}
physeq.active <- physeq.prop.wat
```

### Bray Curtis metric

```{r }
set.seed(1234)

finalPhyseq.ord <- ordinate(physeq.active, "PCoA", "bray")
p1 = plot_ordination(physeq.active,
                     finalPhyseq.ord,
                     type="samples",
                     color="Environment",
                     shape = "Simple_extraction",
                     title="Samples") + geom_point(size = 2.5)
p1 + ggtitle("PCoA, Bray Curtis distance")
#ggplotly(p1 + ggtitle("PCoA, Bray Curtis distance"))
#+  scale_fill_viridis_d()
```

Very clearly splitting by Environment and extraction type. Clustering isn't as tight as it was for larvae (not too surprising).

Let's try highlighting Density and Competition...

```{r }
set.seed(1234)

finalPhyseq.ord <- ordinate(physeq.active, "PCoA", "bray")
p1 = plot_ordination(physeq.active,
                     finalPhyseq.ord,
                     type="samples",
                     color="Environment",
                     shape = "Density",
                     label = "Environment",
                     title="PCoA, Bray Curtis distance") + 
  geom_point(size = 3, aes(fill=Competition)) + 
  scale_shape_manual(values = c(21, 22, 23) )+
  scale_color_manual(values = c("black","gray") ) +
  #scale_fill_manual(values=c("#EC9DAE","#AEDE94", "#9dd6ec")) +
  scale_fill_manual(values=c("black","darkgray", "white")) +
  guides(fill=guide_legend(override.aes=list(shape=21)))+
  guides(color=guide_legend(override.aes=list(shape=21))) 
  #guides(fill=guide_legend(override.aes=list(shape=21,fill=c("#EC9DAE","#AEDE94","#9dd6ec"),color=c("black","white"))))
  guides(fill=guide_legend(override.aes=list(shape=21,fill=c("black","darkgray", "white"),color=c("black","gray"))))
p1
ggsave("pcoa-braycurtis-density-competition-environment_ExpWater_bgw.pdf", path = "../results/final-plots/", device = "pdf")
```

Infusion samples definitely separate (centered in this plot), but everything else seems quite mixed.


Check to see if number of sequences per sample (input) affects diversity

```{r }
p1 <- plot_ordination(physeq.active,
                     finalPhyseq.ord,
                     type="samples",
                     color="input",
                     label="Label",
                     title="Samples") + geom_point(size = 2.5) +
  ggtitle("PCoA, Bray Curtis distance") 
ggplotly(p1)
```

Nothing odd related to number of input sequences.


### Weighted UniFrac metric

Let's look at Weighted UniFrac now.

```{r}
finalPhyseq.ord <- ordinate(physeq.active, "PCoA", "wunifrac")
p1 = plot_ordination(physeq.active,
                     finalPhyseq.ord,
                     type="samples",
                     color="Environment",
                     shape = "Simple_extraction",
                    # label = "Label",
                     title="PCoA, Weighted Unifrac distance") + geom_point(size = 2.5)
p1

```

Very similar pattern as Bray-Curtis. 

Let's also highlight the other metadata of interest:

```{r}
finalPhyseq.ord <- ordinate(physeq.active, "PCoA", "wunifrac")
p1 = plot_ordination(physeq.active,
                     finalPhyseq.ord,
                     type="samples",
                     color="Environment",
                     shape = "Density",
                     label = "Environment",
                     title="PCoA, Weighted Unifrac distance") + 
  geom_point(size = 3, aes(fill=Competition)) + 
  scale_shape_manual(values = c(21, 22, 23) )+
  scale_color_manual(values = c("black","gray") ) +
  scale_fill_manual(values=c("#EC9DAE","#AEDE94", "#9dd6ec")) +
  #scale_fill_manual(values=c("black","darkgray", "white")) +
  guides(fill=guide_legend(override.aes=list(shape=21)))+
  guides(color=guide_legend(override.aes=list(shape=21))) 
  guides(fill=guide_legend(override.aes=list(shape=21,fill=c("#EC9DAE","#AEDE94","#9dd6ec"),color=c("black","white"))))
  #guides(fill=guide_legend(override.aes=list(shape=21,fill=c("black","darkgray", "white"),color=c("black","gray"))))
p1

ggsave("pcoa-wunifrac-density-competition-environment_ExpWater_color.pdf", path = "../results/final-plots/", device = "pdf")
```

Infusion still separates from the others. Everything else quite mixed.




## PERMANOVA

Let's try PERMANOVA on this using the implementation in `vegan` (`adonis2`). This is the most popular and powerful test  recommended for microbial studies. P-value indicates whether the diversity between different communities is statistically significant. R^2 indicates how much of the total variance can be explained by grouping factor.

NOTE, these initial tests include both experimental water and mosquito larvae. Infusion stock samples have been removed for the PERMANOVA analyses.

### Bray-Curtis 

Results below are for the Bray-Curtis metric, but I also post results of Weighted-Unifrac after this. I also list results for complex models as well as using the combined metadata column 

```{r }
# Remove infusion samples
#Normalize
physeq.prop <- transform_sample_counts(physeq.prev, function(x) x/sum(x) )
sample_data(physeq.prop)$SampleSums <- sample_sums(physeq.prev)

physeq.active <- physeq.prop

physeq.active <- prune_samples(sample_data(physeq.active)$Density != "Stock Water", physeq.active)

finalPhyseq.meta <- as(sample_data(physeq.active), "data.frame")
finalPhyseq.prop.dist.bc <- phyloseq::distance(physeq.active, method = "bray")
```


Environment x Competition x Density

```{r}
set.seed(12345)
adonis2(finalPhyseq.prop.dist.bc ~ Environment * Competition * Density,
       data = finalPhyseq.meta)
```

All three are considered significant, however these only contribute to ~25% of the variance (R2). How about the metadata column (Env_Comp_Den) by itself?

```{r}
set.seed(12345)
adonis2(finalPhyseq.prop.dist.bc ~ Env_Comp_Den,
       data = finalPhyseq.meta)
```

Significant.


Extraction x Environment x Competition x Density

```{r}
set.seed(12345)
adonis2(finalPhyseq.prop.dist.bc ~ Simple_extraction * Environment * Competition * Density,
       data = finalPhyseq.meta)
```

All 4 are significant, including some of the interactions. Together these do account for ~69% of the variance (R2), so this seems like a more complete model. Let's try just the metadata column (Extr_Env_Comp_Den).

```{r}
set.seed(12345)
adonis2(finalPhyseq.prop.dist.bc ~ Extr_Env_Comp_Den,
       data = finalPhyseq.meta)
```

Significant.

For good measure, let's also add Plate to the above model.

```{r}
set.seed(12345)
adonis2(finalPhyseq.prop.dist.bc ~ Simple_extraction * Environment * Competition * Density * Plate,
       data = finalPhyseq.meta)
```

Adding plate increased the accounted variance by maybe 1% and it is also significant. Looking at the metadata it does seem that Plate is a confounding variable and is very closely related to extraction type. It's not likely adding much variation in reality, but it can't be disentangled from extraction_type.


#### Heteroscedasicity test
Now let's test for differences between group dispersion. Ideally, these tests won't be significant, as heteroskedasticity can can affect the validity of a PERMANOVA.

Env_Comp_Den

```{r}
beta <- betadisper(finalPhyseq.prop.dist.bc, finalPhyseq.meta$Env_Comp_Den)
permutest(beta)
```

Env_Comp_Den is not heteroscedastic! Great.


Extr_Env_Comp_Den

```{r}
beta <- betadisper(finalPhyseq.prop.dist.bc, finalPhyseq.meta$Extr_Env_Comp_Den)
permutest(beta)
```

Extr_Env_Comp_Den IS heteroscedastic, so this could affect the results of a PERMANOVA. I suspect that extraction type is making the data unbalanced, so we could separate the larvae from the water samples and try this again.



```{r}
plot(beta, main = "beta dispersion estimates between days")
```



### Weighted Unifrac

```{r }
finalPhyseq.meta <- as(sample_data(physeq.active), "data.frame")
finalPhyseq.prop.dist.bc <- phyloseq::distance(physeq.active, method = "wunifrac")
```


Environment x Competition x Density

```{r}
set.seed(12345)
adonis2(finalPhyseq.prop.dist.bc ~ Environment * Competition * Density,
       data = finalPhyseq.meta)
```

All three are considered significant, however these only contribute to ~25% of the variance (R2). How about the metadata column (Env_Comp_Den) by itself?

```{r}
set.seed(12345)
adonis2(finalPhyseq.prop.dist.bc ~ Env_Comp_Den,
       data = finalPhyseq.meta)
```
Significant.


Extraction x Environment x Competition x Density

```{r}
set.seed(12345)
adonis2(finalPhyseq.prop.dist.bc ~ Simple_extraction * Environment * Competition * Density,
       data = finalPhyseq.meta)
```

All 4 are significant, including some of the interactions. Together these do account for ~80% of the variance (R2), a little higher than Bray-Curtis. Let's try just the metadata column (Extr_Env_Comp_Den).

```{r}
set.seed(12345)
adonis2(finalPhyseq.prop.dist.bc ~ Extr_Env_Comp_Den,
       data = finalPhyseq.meta)
```
Significant.

For good measure, let's also add Plate to the above model.

```{r}
set.seed(12345)
adonis2(finalPhyseq.prop.dist.bc ~ Simple_extraction * Environment * Competition * Density * Plate,
       data = finalPhyseq.meta)
```

Everything is significant and accounts for ~83% of the variance (3% more).


#### Heteroscedasicity test
Now let's test for differences between group dispersion. Ideally, these tests won't be significant, as heteroskedasticity can can affect the validity of a PERMANOVA.

Env_Comp_Den

```{r}
beta <- betadisper(finalPhyseq.prop.dist.bc, finalPhyseq.meta$Env_Comp_Den)
permutest(beta)
```

Env_Comp_Den is not heteroscedastic! Great.


Extr_Env_Comp_Den

```{r}
beta <- betadisper(finalPhyseq.prop.dist.bc, finalPhyseq.meta$Extr_Env_Comp_Den)
permutest(beta)
```

Extr_Env_Comp_Den IS less heteroscedastic than the Bray-Curtis results, but still significant. 


Let's visualize

```{r}
plot(beta, main = "beta dispersion estimates between days")
```


### Test larvae only (using Weighted Unifrac)

Here I test larvae only to see if that would make the data more balanced and remove heterscedasticity.

```{r}
physeq.lar <- prune_samples(sample_data(physeq.active)$Simple_extraction == "Mosquito Larvae", physeq.active)
#physeq.wat <- prune_samples(sample_data(physeq.active)$Simple_extraction == "Water", physeq.active)
```

```{r }
finalPhyseq.meta <- as(sample_data(physeq.lar), "data.frame")
finalPhyseq.prop.dist.bc <- phyloseq::distance(physeq.lar, method = "wunifrac")
```


Environment x Competition x Density

```{r}
set.seed(12345)
adonis2(finalPhyseq.prop.dist.bc ~ Environment * Competition * Density,
       data = finalPhyseq.meta)
```

All 3 are significant and this does increase R2 is around 43%

Environment x Competition x Density x Plate

```{r}
set.seed(12345)
adonis2(finalPhyseq.prop.dist.bc ~ Environment * Competition * Density * Plate,
       data = finalPhyseq.meta)
```

Adding Plate increases significance for all 4, and increases R2 to ~59%. 

Heteroscedasticity

```{r}
beta <- betadisper(finalPhyseq.prop.dist.bc, finalPhyseq.meta$Env_Comp_Den)
permutest(beta)
```

Env_Comp_Den is not heteroscedastic.

Let's visualize

```{r}
plot(beta, main = "beta dispersion estimates between days")
```



### Test larvae only (using Bray Curtis)

Here I wanted to test larvae only to see if that would make the data more balanced and remove heterscedasticity.


```{r }
finalPhyseq.meta <- as(sample_data(physeq.lar), "data.frame")
finalPhyseq.prop.dist.bc <- phyloseq::distance(physeq.lar, method = "bray")
```


Environment x Competition x Density

```{r}
set.seed(12345)
adonis2(finalPhyseq.prop.dist.bc ~ Environment * Competition * Density,
       data = finalPhyseq.meta)
```

All 3 are significant and this does increase R2 is around 55%

Environment x Competition x Density x Plate

```{r}
set.seed(12345)
adonis2(finalPhyseq.prop.dist.bc ~ Environment * Competition * Density * Plate,
       data = finalPhyseq.meta)
```

Adding Plate increases significance for all 4, and increases R2 to ~60%. 

Heteroscedasticity

```{r}
beta <- betadisper(finalPhyseq.prop.dist.bc, finalPhyseq.meta$Env_Comp_Den)
permutest(beta)
```

Env_Comp_Den is not heteroscedastic.

Let's visualize

```{r}
plot(beta, main = "beta dispersion estimates between days")
```