```{r, include=FALSE}
source("~/Desktop/hpcbio-git/caceres-16S-mosquito-2022Sept/src/common_code.R", local = knitr::knit_global())
```

```{r}
physeq.prev <- readRDS('results/phyloseq.prevfiltered.RDS')
```

# Differential abundance analysis

This abundance testing should help us see the differences between water samples and larvae samples. Will start with comparing those two, but I can do other comparisons as well.

Run DESeq2. Will use DESeq2's normalization method since it is more appropriate for abundance testing. 

```{r echo=FALSE, message=FALSE}
library(DESeq2)
sd <- sample_data(physeq.prev)

sd$Group <- sd$Simple_extraction
sample_data(physeq.prev) <- sd

design <- model.matrix(~ 0 + Group, data = as(sd, "data.frame"))

finalPhyseq.adds = phyloseq_to_deseq2(physeq.prev, ~ 0 + Group)

finalPhyseq.adds = DESeq(finalPhyseq.adds,
                         test="Wald",
                         fitType="local",
                         sfType = "poscounts",
                         quiet = TRUE
                         )
```

What are the results names?

```{r}
resultsNames(finalPhyseq.adds)
```

```{r }
generateContrastResults <- function(deseq, contrast, physeq, file = "results.txt") {
  res = results(deseq, 
              cooksCutoff = FALSE, 
              contrast = contrast,
              test = "Wald"
              )
  res = cbind(as(res, "data.frame"), as(tax_table(physeq)[rownames(res), ], "matrix"))
  
  write.table(res, file, row.names = TRUE, col.names = NA, sep = "\t")
  return(res)
}

plotDESeq2ResGenus <- function(res) {
  sigtab <- res[which(res$padj < 0.05), ]
  scale_fill_discrete <- function(palname = "Set1", ...) {
      scale_fill_brewer(palette = palname, ...)
  }
  
  x = tapply(sigtab$log2FoldChange, sigtab$Phylum, function(x) max(x))
  x = sort(x, TRUE)
  sigtab$Phylum = factor(as.character(sigtab$Phylum), levels=names(x))
  
  # Genus order
  x = tapply(sigtab$log2FoldChange, sigtab$Genus, function(x) max(x))
  x = sort(x, TRUE)
  sigtab$Genus = factor(as.character(sigtab$Genus), levels=names(x))
  p <- ggplot(sigtab, aes(x=Genus, y=log2FoldChange, color=Phylum)) + geom_point(size=6) +
    theme(axis.text.x = element_text(angle = -45, hjust = 0, vjust=0.5, face = 'italic'),
          legend.text = element_text(face = 'italic'))
  return(p)
}

plotDESeq2ResFamily <- function(res) {
  sigtab <- res[which(res$padj < 0.05), ]
  scale_fill_discrete <- function(palname = "Set1", ...) {
      scale_fill_brewer(palette = palname, ...)
  }
  
  x = tapply(sigtab$log2FoldChange, sigtab$Phylum, function(x) max(x))
  x = sort(x, TRUE)
  sigtab$Phylum = factor(as.character(sigtab$Phylum), levels=names(x))
  
  # Family order
  x = tapply(sigtab$log2FoldChange, sigtab$Family, function(x) max(x))
  x = sort(x, TRUE)
  sigtab$Family = factor(as.character(sigtab$Family), levels=names(x))
  p <- ggplot(sigtab, aes(x=Family, y=log2FoldChange, color=Phylum)) + geom_point(size=6) +
    theme(axis.text.x = element_text(angle = -45, hjust = 0, vjust=0.5, face = 'italic'),
          legend.text = element_text(face = 'italic'))
  return(p)
}

plotDESeq2ResSpecies <- function(res) {
  sigtab <- res[which(res$padj < 0.05), ]
  scale_fill_discrete <- function(palname = "Set1", ...) {
      scale_fill_brewer(palette = palname, ...)
  }
  
  x = tapply(sigtab$log2FoldChange, sigtab$Phylum, function(x) max(x))
  x = sort(x, TRUE)
  sigtab$Phylum = factor(as.character(sigtab$Phylum), levels=names(x))
  
  # Genus + Species order
  sigtab$GeneSpecies <- paste(sigtab$Genus,sigtab$Species)
  x = tapply(sigtab$log2FoldChange, sigtab$GeneSpecies, function(x) max(x))
  x = sort(x, TRUE)
  sigtab$GeneSpecies = factor(as.character(sigtab$GeneSpecies), levels=names(x))
  p <- ggplot(sigtab, aes(x=GeneSpecies, y=log2FoldChange, color=Phylum)) + geom_point(size=6) +
    theme(axis.text.x = element_text(angle = -45, hjust = 0, vjust=0.5, face = 'italic'),
          legend.text = element_text(face = 'italic'))
  return(p)
}

```

Pull out specific contrast. The numbers in the contrast vector below correspond to the different test results we generated:

## Water vs Larvae

This is just an example of the types of plots I can make with differential abundance testing. Let me know if you would like me to pull out other or all comparisons. I will save them to pdfs rather than putting them in this report.

How many taxa are significant?

```{r }
res.test = generateContrastResults(deseq = finalPhyseq.adds,
                              contrast = c("Group", "Water", "Mosquito Larvae"),
                              physeq = physeq.prev,
                              file = "results/DESeq2/Overall-Water-vs-Larvae.txt"
)
dim(res.test[which(res.test$padj <= 0.05), ])
```

There are 115 significant ASVs.


Plot at Family-level
```{r }
plotDESeq2ResFamily(res.test)
ggsave("Diff-Abund_Family_Water-vs-Larvae.pdf", path = "results/DESeq2", width = 11, height = 8,device = "pdf")
```

Plot at Genus-level
```{r }
plotDESeq2ResGenus(res.test)
ggsave("Diff-Abund_Genus_Water-vs-Larvae.pdf", path = "results/DESeq2/", width = 12, height = 8, device = "pdf")
```

Plot at Species-level
```{r }
plotDESeq2ResSpecies(res.test)
ggsave("Diff-Abund_Species_Water-vs-Larvae.pdf", path = "results/DESeq2/", width = 18, height = 10, device = "pdf")
```

## Stock Water vs Water

```{r }
res.test = generateContrastResults(deseq = finalPhyseq.adds,
                              contrast = c("Group", "Stock Water", "Water"),
                              physeq = physeq.prev,
                              file = "results/DESeq2/Overall-StockWater-vs-Water.txt"
)
dim(res.test[which(res.test$padj <= 0.05), ])
```

There are 74 significant ASVs.

Plot at Family-level
```{r }
plotDESeq2ResFamily(res.test)
ggsave("Diff-Abund_Family_Stock.Water-vs-Water.pdf", path = "results/DESeq2", width = 8, height = 6,device = "pdf")
```

Plot at Genus-level
```{r }
plotDESeq2ResGenus(res.test)
ggsave("Diff-Abund_Genus_Stock.Water-vs-Water.pdf", path = "results/DESeq2/", width = 10, height = 6, device = "pdf")
```

Plot at Species-level
```{r }
plotDESeq2ResSpecies(res.test)
ggsave("Diff-Abund_Species_Stock.Water-vs-Water.pdf", path = "results/DESeq2/", width = 12, height = 6, device = "pdf")
```

## Density 20 vs 40 (Larvae Only)

```{r }
# Pull out Larvae only
physeq.lar <- prune_samples(sample_data(physeq.prev)$Simple_extraction == "Mosquito Larvae", physeq.prev)

sd <- sample_data(physeq.lar)

sd$Group <- sd$Density
sample_data(physeq.lar) <- sd

design <- model.matrix(~ 0 + Group, data = as(sd, "data.frame"))

finalPhyseq.adds = phyloseq_to_deseq2(physeq.lar, ~ 0 + Group)

finalPhyseq.adds = DESeq(finalPhyseq.adds,
                         test="Wald",
                         fitType="local",
                         sfType = "poscounts",
                         quiet = TRUE
                         )
```

What are the results names?

```{r}
resultsNames(finalPhyseq.adds)
```

How many taxa are significant?

```{r }
res.test = generateContrastResults(deseq = finalPhyseq.adds,
                              contrast = c("Group", "20", "40"),
                              physeq = physeq.prev,
                              file = "results/DESeq2/Overall-20-vs-40.txt"
)
dim(res.test[which(res.test$padj <= 0.05), ])
```

There are 12 significant ASVs.


Plot at Family-level
```{r }
plotDESeq2ResFamily(res.test)
ggsave("Diff-Abund_Family_20-vs-40.pdf", path = "results/DESeq2", width = 5, height = 4.5, device = "pdf")
```

Plot at Genus-level
```{r }
plotDESeq2ResGenus(res.test)
ggsave("Diff-Abund_Genus_20-vs-40.pdf", path = "results/DESeq2/", width = 5, height = 4.5, device = "pdf")
```

Plot at Species-level
```{r }
plotDESeq2ResSpecies(res.test)
ggsave("Diff-Abund_Species_20-vs-40.pdf", path = "results/DESeq2/", width = 5, height = 5.5, device = "pdf")
```



## Environment G vs GN (Larvae Only)

```{r }

sd <- sample_data(physeq.lar)

sd$Group <- sd$Environment
sample_data(physeq.lar) <- sd

design <- model.matrix(~ 0 + Group, data = as(sd, "data.frame"))

finalPhyseq.adds = phyloseq_to_deseq2(physeq.lar, ~ 0 + Group)

finalPhyseq.adds = DESeq(finalPhyseq.adds,
                         test="Wald",
                         fitType="local",
                         sfType = "poscounts",
                         quiet = TRUE
                         )
```

What are the results names?

```{r}
resultsNames(finalPhyseq.adds)
```

How many taxa are significant?

```{r }
res.test = generateContrastResults(deseq = finalPhyseq.adds,
                              contrast = c("Group", "G", "GN"),
                              physeq = physeq.prev,
                              file = "results/DESeq2/Overall-G-vs-GN.txt"
)
dim(res.test[which(res.test$padj <= 0.05), ])
```

There are 109 significant ASVs.


Plot at Family-level
```{r }
plotDESeq2ResFamily(res.test)
ggsave("Diff-Abund_Family_G-vs-GN.pdf", path = "results/DESeq2", width = 10, height = 5,device = "pdf")
```

Plot at Genus-level
```{r }
plotDESeq2ResGenus(res.test)
ggsave("Diff-Abund_Genus_G-vs-GN.pdf", path = "results/DESeq2/", width = 14, height = 6, device = "pdf")
```

Plot at Species-level
```{r }
plotDESeq2ResSpecies(res.test)
ggsave("Diff-Abund_Species_G-vs-GN.pdf", path = "results/DESeq2/", width = 18, height = 8, device = "pdf")
```





<!-- ## Create all Diff Abund plots & save to PDF -->

<!-- ```{r } -->
<!-- library(DESeq2) -->
<!-- sd <- sample_data(physeq.prev) -->

<!-- sd$Group <- sd$Treatment_Day -->
<!-- sample_data(physeq.prev) <- sd -->

<!-- design <- model.matrix(~ 0 + Group, data = as(sd, "data.frame")) -->

<!-- finalPhyseq.adds = phyloseq_to_deseq2(physeq.prev, ~ Group) -->

<!-- finalPhyseq.adds = DESeq(finalPhyseq.adds, -->
<!--                          test="Wald", -->
<!--                          fitType="local", -->
<!--                          sfType = "poscounts", -->
<!--                          quiet = TRUE -->
<!--                          ) -->
<!-- ``` -->

<!-- What are the results names? -->

<!-- ```{r} -->
<!-- resultsNames(finalPhyseq.adds) -->
<!-- ``` -->

<!-- Run a loop that will save these tables and images for all the different pairwise contrasts. -->

<!-- Treatment x Day  -->
<!-- ```{r} -->
<!-- n.all <- sample_data(physeq.prev)$Treatment_Day %>% levels() -->
<!-- for(x in 1:16){ -->
<!--   for(y in 1:16){ -->
<!--     if(x < y){  # Ignore useless and redundant files -->
<!--       res.test <- generateContrastResults(deseq = finalPhyseq.adds,  -->
<!--                                 contrast = c("Group",  -->
<!--                                              n.all[x],  # num -->
<!--                                              n.all[y]), # denom -->
<!--                                 physeq = physeq.prev, -->
<!--                                 file = paste("results/diff-abund/Treatment_Day/tables/Diff-Abund_", n.all[x], "-vs-", n.all[y], ".txt", sep="") -->
<!--       ) -->
<!--       plotDESeq2ResFamily(res.test) -->
<!--       f <- paste("Diff-Abund_Family_", n.all[x], "-vs-", n.all[y], ".pdf", sep="") -->
<!--       ggsave(f, path = "results/diff-abund/Treatment_Day", device = "pdf") -->
<!--       plotDESeq2ResGenus(res.test) -->
<!--       f <- paste("Diff-Abund_Genus_", n.all[x], "-vs-", n.all[y], ".pdf", sep="") -->
<!--       ggsave(f, path = "results/diff-abund/Treatment_Day", device = "pdf") -->
<!--       plotDESeq2ResSpecies(res.test) -->
<!--       f <- paste("Diff-Abund_Species_", n.all[x], "-vs-", n.all[y], ".pdf", sep="") -->
<!--       ggsave(f, path = "results/diff-abund/Treatment_Day", device = "pdf") -->
<!--     } -->
<!--   } -->
<!-- } -->
<!-- ``` -->


<!-- What about differential abundance for Treatment only? -->

<!-- ```{r } -->
<!-- sd <- sample_data(physeq.prev) -->

<!-- sd$Group <- sd$Treatment -->
<!-- sample_data(physeq.prev) <- sd -->

<!-- design <- model.matrix(~ Group, data = as(sd, "data.frame")) -->

<!-- finalPhyseq.adds = phyloseq_to_deseq2(physeq.prev, ~ Group) -->

<!-- finalPhyseq.adds = DESeq(finalPhyseq.adds, -->
<!--                          test="Wald", -->
<!--                          fitType="local", -->
<!--                          sfType = "poscounts", -->
<!--                          quiet = TRUE -->
<!--                          ) -->
<!-- ``` -->

<!-- What are the results names? -->

<!-- ```{r} -->
<!-- resultsNames(finalPhyseq.adds) -->
<!-- ``` -->

<!-- Run a loop that will save these tables and images for all the different pairwise contrasts. -->
<!-- ```{r} -->
<!-- n.all <- sample_data(physeq.prev)$Treatment %>% levels() -->
<!-- for(x in 1:4){ -->
<!--   for(y in 1:4){ -->
<!--     if(x < y){  # Ignore useless and redundant files -->
<!--       res.test <- generateContrastResults(deseq = finalPhyseq.adds,  -->
<!--                                 contrast = c("Group",  -->
<!--                                              n.all[x],  # num -->
<!--                                              n.all[y]), # denom -->
<!--                                 physeq = physeq.prev, -->
<!--                                 file = paste("results/diff-abund/Treatment/tables/Diff-Abund_", n.all[x], "-vs-", n.all[y], ".txt", sep="") -->
<!--       ) -->
<!--       plotDESeq2ResFamily(res.test) -->
<!--       f <- paste("Diff-Abund_Family_", n.all[x], "-vs-", n.all[y], ".pdf", sep="") -->
<!--       ggsave(f, path = "results/diff-abund/Treatment", device = "pdf") -->
<!--       plotDESeq2ResGenus(res.test) -->
<!--       f <- paste("Diff-Abund_Genus_", n.all[x], "-vs-", n.all[y], ".pdf", sep="") -->
<!--       ggsave(f, path = "results/diff-abund/Treatment", device = "pdf") -->
<!--       plotDESeq2ResSpecies(res.test) -->
<!--       f <- paste("Diff-Abund_Species_", n.all[x], "-vs-", n.all[y], ".pdf", sep="") -->
<!--       ggsave(f, path = "results/diff-abund/Treatment", device = "pdf") -->
<!--     } -->
<!--   } -->
<!-- } -->
<!-- ``` -->

<!-- What about differential abundance for Day only? -->

<!-- ```{r } -->
<!-- sd <- sample_data(physeq.prev) -->

<!-- sd$Group <- sd$Day -->
<!-- sample_data(physeq.prev) <- sd -->

<!-- design <- model.matrix(~ Group, data = as(sd, "data.frame")) -->

<!-- finalPhyseq.adds = phyloseq_to_deseq2(physeq.prev, ~ Group) -->

<!-- finalPhyseq.adds = DESeq(finalPhyseq.adds, -->
<!--                          test="Wald", -->
<!--                          fitType="local", -->
<!--                          sfType = "poscounts", -->
<!--                          quiet = TRUE -->
<!--                          ) -->
<!-- ``` -->

<!-- What are the results names? -->

<!-- ```{r} -->
<!-- resultsNames(finalPhyseq.adds) -->
<!-- ``` -->

<!-- Run a loop that will save these tables and images for all the different pairwise contrasts. -->
<!-- ```{r} -->
<!-- n.all <- sample_data(physeq.prev)$Day %>% levels() -->
<!-- for(x in 1:4){ -->
<!--   for(y in 1:4){ -->
<!--     if(x < y){  # Ignore useless and redundant files -->
<!--       res.test <- generateContrastResults(deseq = finalPhyseq.adds,  -->
<!--                                 contrast = c("Group",  -->
<!--                                              n.all[x],  # num -->
<!--                                              n.all[y]), # denom -->
<!--                                 physeq = physeq.prev, -->
<!--                                 file = paste("results/diff-abund/Day/tables/Diff-Abund_", n.all[x], "-vs-", n.all[y], ".txt", sep="") -->
<!--       ) -->
<!--       plotDESeq2ResFamily(res.test) -->
<!--       f <- paste("Diff-Abund_Family_", n.all[x], "-vs-", n.all[y], ".pdf", sep="") -->
<!--       ggsave(f, path = "results/diff-abund/Day", device = "pdf") -->
<!--       plotDESeq2ResGenus(res.test) -->
<!--       f <- paste("Diff-Abund_Genus_", n.all[x], "-vs-", n.all[y], ".pdf", sep="") -->
<!--       ggsave(f, path = "results/diff-abund/Day", device = "pdf") -->
<!--       plotDESeq2ResSpecies(res.test) -->
<!--       f <- paste("Diff-Abund_Species_", n.all[x], "-vs-", n.all[y], ".pdf", sep="") -->
<!--       ggsave(f, path = "results/diff-abund/Day", device = "pdf") -->
<!--     } -->
<!--   } -->
<!-- } -->
<!-- ``` -->