# Initial filtering

```{r, include=FALSE}
# at some point this should be moved into bookdown
source("~/Desktop/hpcbio-git/caceres-16S-mosquito-2022Sept/src/common_code.R", local = knitr::knit_global())
```

```{r}
# Load in data
physeq <- readRDS('results/physeq.raw.RDS')
physeq.filtered <- physeq
```

These steps remove or subset the original data by:

1. Selecting samples from the original to focus on
2. Remove any possible bad/uninformative samples (with justification)
3. Remove any taxa not found in the remaining samples after removing the above
4. Remove artifacts
5. Remove contaminants from cross-contamination (requires controls) 
6. Prune out low-count taxa

We save this file for alpha rarefaction, alpha diversity, and prevalence filtering/agglomeration

## Subset of data

Are we focusing on one particular subset of samples?  

Not at the moment, so will skip the following step.

### Subset samples


```{r}
#physeq.filtered <- subset_samples(physeq, Sex == "M")
#physeq.filtered
```

## Prune out any other samples?

Will remove the sample that has less than 5000 counts.

```{r }
#physeq.filtered <- prune_samples(!(sample_data(physeq.filtered)$Density == "Infusion"), physeq.filtered)
physeq.filtered <- prune_samples(sample_sums(physeq.filtered) > 10000, physeq.filtered)
physeq.filtered
```

## Zero-count taxa

If we remove samples, there may be some taxa that are no longer present in the subset we have.  Any taxa present that are 0 count?

```{r}
table(taxa_sums(physeq.filtered) == 0)
```

After removing Infusion samples, 579 are 0 count.

```{r}
physeq.filtered <- subset_taxa(physeq.filtered, taxa_sums(physeq.filtered) > 0)
physeq.filtered
```

If needed, save these as a separate input for the next steps.

```{r}
#saveRDS(physeq.filtered, 'results/physeq.raw.males.RDS')
```

## Removing artifacts

We need to remove ASVs that are likely artefactual, for example from host or from other sources.  We should get rid of:

* Unranked/unassigned
* Eukarya (kingdom)
* Mitochondria
* Chloroplast

The RDP classifier and other classifiers will sometimes misclassify sequences that have very little sequence similarity to 16S based on the way it works; these tend to classify poorly by rank (NA or 'Unclassified' up to and sometimes including Domain/Kingdom).  

What proportion of the ASVs are unassigned at each rank?

```{r}
taxtab <- tax_table(physeq.filtered)

# note this test checks if the rank is NA; if the rank is assigned 'Unclassified' modify the term here
ranks <- data.frame(apply(taxtab, 2, 
                          function(x) label_percent(accuracy = 0.01) (sum( is.na(x) )/length(x) ) ))
colnames(ranks) <- c("Perc.Unclassified")
ranks %>% kable
```

How many Domain/Kingdom levels are there?

```{r }
table(tax_table(physeq.filtered)[,'Domain']) %>% kable
```

There aren't any non-microbial hits, so no need to remove sequences mis-classified as somehow eukaryotic

There are a few that are unclassified across almost all ranks; anything not classified at least to a phylum level are uninformative.  We should go ahead and remove these. 

How many Phylum are unassigned?  (TRUE means these are unassigned)

```{r }
table(is.na(tax_table(physeq.filtered)[,'Phylum'])) %>% kable
```

14 are not assigned at the phylum rank.  Export the unassigned sequences to a file.

```{r}
unasn <- subset_taxa(physeq.filtered, is.na(Phylum))
unasn
```

```{r}
library(Biostrings)
dir.create(file.path('..', 'results', 'unasn-sequences'), showWarnings = FALSE, recursive = TRUE)
writeXStringSet(refseq(unasn), 
                filepath = '../results/unasn-sequences/unassigned-phylum-level.fna', 
                format = 'fasta')
```

Subset those sequences that have assignments at least at the Phylum level.

```{r }
physeq.filtered <- subset_taxa(physeq.filtered, !(is.na(Phylum)))
physeq.filtered
```

### Eukaryotic sequences

None were found that were explicitly classified as eukarya, so skipping.

<!-- ```{r } -->
<!-- # extract eukaryotic sequences if found -->
<!-- euks <- subset_taxa(physeq.pacbio, Domain == "Eukaryota") -->
<!-- tax_table(euks) %>% as.data.frame %>% knitr::kable() -->
<!-- ``` -->

<!-- ```{r } -->
<!-- tmp <- sample_sums(euks) -->
<!-- tmp[ tmp > 0 ] -->
<!-- ``` -->

<!-- ```{r } -->
<!-- physeq.pacbio <- subset_taxa(physeq.pacbio, !(Domain == 'Eukaryota')) -->
<!-- ``` -->

### Organelle 16S (mitochondria, chloroplast)

Next we check for organelle 16S.  These could arise in chloroplast or mitochondria.  They are classified under different rank criteria in Silva, so we need to check for them separately. 

Any hits for chloroplast? These are found under the `Class` rank.

```{r }
table(tax_table(physeq.filtered)[,'Order'])['Chloroplast']
```
No chloroplast sequences found.

```{r}
# chloros <- subset_taxa(physeq.filtered, Order == 'Chloroplast')
# sample_sums(chloros) %>% magrittr::extract(. > 0)
```


Any mitochondrial sequences?  

```{r }
table(tax_table(physeq.filtered)[,'Family'])['Mitochondria']
```

No mitochrondial sequence found.

```{r }
#mitos <- subset_taxa(physeq.filtered, Family == 'Mitochondria')
#sample_sums(mitos) %>% magrittr::extract(. > 0)
```


```{r }
# remove negative control samples that aren't part of the analysis (water, Zymo)
#physeq.filtered <- subset_samples(physeq, Treatment != 'Control')

# remove mitochondria
#physeq.filtered <- subset_taxa(physeq.filtered, is.na(Family) | Family != 'Mitochondria')

# remove chloroplast
#physeq.filtered <- subset_taxa(physeq.filtered, is.na(Order) | Order != 'Chloroplast')

#physeq.filtered
```


## Remove taxa with 0 counts

Do we need to filter out any taxa with 0 counts?  Let's check how many have taxonomic sums equal to 0

```{r }
table(taxa_sums(physeq.filtered) == 0)
```

None!  

<!-- Removing these below. -->

<!-- ```{r } -->
<!-- physeq.filtered <- prune_taxa(taxa_sums(physeq.filtered) > 0, physeq.filtered) -->
<!-- physeq.filtered -->
<!-- ``` -->

How about the number of samples who have 0 counts?

```{r }
table(sample_sums(physeq.filtered) == 0)
```

No samples with zero counts after filtering in the above steps. No filtering needed.

```{r }
# sample_data(physeq)[sample_sums(physeq.filtered) == 0]
```

```{r }
# physeq.filtered <- prune_samples(sample_sums(physeq.filtered) > 0, physeq)
# physeq.filtered
```

## Contaminant removal from control samples

We do not have background controls in this experiment therefore can't do this step.  

```{r }
# These steps use the decontam library

# df.contam <- as.data.frame(sample_data(physeq.filtered))
# df.contam$LibrarySize <- sample_sums(physeq.filtered)

# sample_data(physeq.filtered)$is.neg <- sample_data(physeq.filtered)$Control == "Negative"
# contamdf.prev <- isContaminant(physeq.filtered, 
#                                method = "prevalence", 
#                                neg = "is.neg", 
#                                threshold = 0.1)
# 
# table(contamdf.prev$contaminant)
```

<!-- Which ones are contaminants? -->

```{r }
# which(contamdf.prev$contaminant)
```

<!-- Based on the threshold we are seeing some contaminants showing up; how prevalent are they in the overall samples? -->

```{r }
# ps.neg <- prune_samples(
#   sample_data(physeq.filtered)$Control == "Negative", physeq.filtered)
# ps.neg.presence <- transform_sample_counts(ps.neg, function(abund) 1*(abund > 0))
# 
# ps.pos <- prune_samples(
#   sample_data(physeq.filtered)$Control == "Sample", physeq.filtered)
# ps.pos.presence <- transform_sample_counts(ps.pos, function(abund) 1*(abund > 0))
# 
# df.pres <- data.frame(
#   prevalence.pos = taxa_sums(ps.pos.presence), 
#   prevalence.neg = taxa_sums(ps.neg.presence),
#   contam.prev = contamdf.prev$contaminant)
# 
# ggplot(data = df.pres, aes(x=prevalence.neg, y=prevalence.pos, color=contam.prev)) + 
#   geom_point()
```

```{r }
# physeq.filtered <- prune_taxa(!contamdf.prev$contaminant, physeq.filtered)
# physeq.filtered
```

## Prune low count samples

We skip this in favor of precedence filtering, but will plot a histogram of sample counts to show the overall distribution.

```{r}
hist(sample_sums(physeq.filtered), breaks = 50)
```

Pretty reasonable spread.


## Basic overview

Let's look at the data.  We have `r ntaxa(physeq.filtered)` taxa and `r nsamples(physeq.filtered)` samples.

Here are the sample variables for analysis:

```{r }
sample_variables(physeq.filtered)
```

Note these include data from read tracking.  Here is a simple summary of the phylogenetic tree (this is midpoint-rooted in the workflow) with the top 50 most abundant ASVs.

```{r }
# get the top 50 taxa based on overall taxa sums
myTaxa1 <- names(sort(taxa_sums(physeq.filtered), decreasing = TRUE))[1:50]

# generate a tmp instance with only those taxa
ex0 = prune_taxa(myTaxa1, physeq.filtered)

# plot with only those, labeling as needed
plot_tree(ex0, label.tips = 'Family',
          ladderize = "left",
          justify = "left",
          color = 'Simple_extraction'
          )
```

We can see that the 50 most abundant ASVs are dominant in water samples, and seeing a lot of Arcobacteraceae family in all extraction types.

```{r}
saveRDS(physeq.filtered, 'results/physeq.filtered.RDS')
```
