```{r, include=FALSE}
# at some point this should be moved into bookdown
source("~/Desktop/hpcbio-git/caceres-16S-mosquito-2022Sept/src/common_code.R", local = knitr::knit_global())
```

```{r}
physeq <- readRDS('results/physeq.raw.RDS')
```

# Composition plots (unfiltered data)

Overall compositional summary (stacked bar plots) for all unfiltered samples.  In this example I pulled out the top 20 families by composition per sample, split by treatment. Data has been transformed to relative abundance (no prior added).

There are no Species-level assignments for top 20 families, and the top 20 families only make up, at most, 20% of the total abundance. This indicates that we have very diverse communities. This is not surprising considering that we have 18,538 ASVs to start with. It will be beneficial to glom ASVs at the species or genus level to reduce the number of ASVs to something more manageable. But I don't imagine these raw composition plots will be particularily useful until after prevalence filtering and taxa glomming.

```{r }
#Get relative abundance
physeq.prop <- transform_sample_counts(physeq, function(x) x/sum(x) )
sample_data(physeq.prop)$SampleSums <- sample_sums(physeq)

#Pull out top 20 ASVs
#physeq.subset <- names(sort(taxa_sums(physeq), decreasing=TRUE))[1:20]
#physeq.plot <- prune_taxa(physeq.subset, physeq.prop)

#Pull out top 20 families
physeq.subset <- tax_glom(physeq, taxrank = "Family")
tax.ordered <- names(sort(taxa_sums(physeq), TRUE)[1:20])
physeq.plot <- subset_taxa(physeq.prop, taxa_names(physeq) %in% tax.ordered)
```




```{r, results='asis', fig.keep='all', message = FALSE, warning = FALSE, echo = FALSE}
#kc <- c("SampleID","Label","Env_Comp_Den", "Replicate")
ranks <- c("Class", "Order", "Family", "Genus")
plots <- lapply(ranks, function (x) { 

  #Plot relative abundance
  p <- plot_bar(physeq.plot, x="Label", fill = x)
  p1 <- p + theme(legend.key.size = unit(0.25, 'cm'), #change legend key size
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=8),
        axis.title.x=element_blank(),
        axis.text.x = element_text(size = 5),
        axis.ticks.x = element_blank()) +
        ylab("Relative Abundance") +
        facet_wrap(~Env_Comp_Den, scales = "free_x")
  
#p <- composition_df(physeq, x, minprop = 0.05, keepcols = kc)
#p + facet_wrap(~Location, scales = "free_x", nrow = 1)
  #ggplot(p, aes(x = Label, y = Proportion, fill = print(x, quote=FALSE))) + geom_col() + 
  #  facet_wrap(~Treatment, scales = "free_x") + 
  #  scale_fill_manual(values = dittoSeq::dittoColors(1)) +
  #  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1, vjust = 0.5))
 ggplotly(p1)
})

names(plots) <- ranks

# Build list of outputs
# See https://stackoverflow.com/questions/9469504/access-and-preserve-list-names-in-lapply-function for example
# not sure if there is an easy way to make a function out of these.
output <- list()
for(rank in ranks){
  # Header for iteration, note Rmd heading ranks and adjust accordingly
  output[[length(output) + 1L]] <- paste0("## ", rank)

  # Plot
  output[[length(output) + 1L]] <- plots[[rank]]
}

# Render the outputs
for(j in 1:length(output)){
  x <- output[[j]]

  if(inherits(x, "character")){
    cat("\n")
    cat(x)
  } else if(inherits(x, "knitr_kable")){
    cat("\n")
    print(x)
  }
  else {
    # print the html piece of the htmlwidgets
    cat("\n")
    cat(htmltools::renderTags(as_widget(x))$html)
  }
}
```

```{r echo=FALSE, messages=FALSE, warning=FALSE}
# Attach the Dependencies since they do not get included with renderTags(...)$html
deps <- lapply(
  Filter(f = function(x){inherits(x,"htmlwidget")}, x = output),
  function(hw){
    htmltools::renderTags(hw)$dependencies
  }
)
htmltools::attachDependencies(x = htmltools::tagList(), value = unlist(deps,recursive=FALSE))
```
