```{r, include=FALSE}
# at some point this should be moved into bookdown
source("~/Desktop/hpcbio-git/hergenrother-16S-mouse-2022Sept/src/common_code.R", local = knitr::knit_global())
```

```{r}
physeq <- readRDS('results/physeq.raw.RDS')
```

# Composition plots (unfiltered data)

Overall compositional summary (stacked bar plots) for all unfiltered samples.  In this example here are the top 20 families by composition per sample, split by treatment. Data has been transformed to relative abundance (no prior added). Note again that most can be classified to the family or order rank, but genus and species ranks see few reliable assignments. 


```{r }
#Get relative abundance
physeq.prop <- transform_sample_counts(physeq, function(x) x/sum(x) )
sample_data(physeq.prop)$SampleSums <- sample_sums(physeq)

#Pull out top 20 families
physeq.subset <- tax_glom(physeq.prop, taxrank = "Family")
tax.ordered <- names(sort(taxa_sums(physeq.prop), TRUE)[1:20])
physeq.plot <- subset_taxa(physeq.prop, taxa_names(physeq.prop) %in% tax.ordered)
```

```{r, results='asis', fig.keep='all', message = FALSE, warning = FALSE, echo = FALSE}
kc <- c("SampleID","Label","Treatment","Day","Replicate")
ranks <- c("Class", "Order", "Family", "Genus", "Species")
plots <- lapply(ranks, function (x) { 

  #Plot relative abundance
  p <- plot_bar(physeq.plot, x="Label", fill = x)
  p1 <- p + theme(legend.key.size = unit(0.25, 'cm'), #change legend key size
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=8),
        axis.title.x=element_blank(),
        axis.text.x = element_text(size = 5),
        axis.ticks.x = element_blank()) +
        ylab("Relative Abundance") +
        facet_wrap(~Treatment, scales = "free_x")
  
#p <- composition_df(physeq, x, minprop = 0.05, keepcols = kc)
#p + facet_wrap(~Location, scales = "free_x", nrow = 1)
  #ggplot(p, aes(x = Label, y = Proportion, fill = print(x, quote=FALSE))) + geom_col() + 
  #  facet_wrap(~Treatment, scales = "free_x") + 
  #  scale_fill_manual(values = dittoSeq::dittoColors(1)) +
  #  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1, vjust = 0.5))
 ggplotly(p1)
})

names(plots) <- ranks

# Build list of outputs
# See https://stackoverflow.com/questions/9469504/access-and-preserve-list-names-in-lapply-function for example
# not sure if there is an easy way to make a function out of these.
output <- list()
for(rank in ranks){
  # Header for iteration, note Rmd heading ranks and adjust accordingly
  output[[length(output) + 1L]] <- paste0("## ", rank)

  # Plot
  output[[length(output) + 1L]] <- plots[[rank]]
}

# Render the outputs
for(j in 1:length(output)){
  x <- output[[j]]

  if(inherits(x, "character")){
    cat("\n")
    cat(x)
  } else if(inherits(x, "knitr_kable")){
    cat("\n")
    print(x)
  }
  else {
    # print the html piece of the htmlwidgets
    cat("\n")
    cat(htmltools::renderTags(as_widget(x))$html)
  }
}
```

```{r echo=FALSE, messages=FALSE, warning=FALSE}
# Attach the Dependencies since they do not get included with renderTags(...)$html
deps <- lapply(
  Filter(f = function(x){inherits(x,"htmlwidget")}, x = output),
  function(hw){
    htmltools::renderTags(hw)$dependencies
  }
)
htmltools::attachDependencies(x = htmltools::tagList(), value = unlist(deps,recursive=FALSE))
```
